Shifts in water supply and demand drive land cover change across Chile

### 2025-01-12 article/article_scientific_reports/mansucript_11012025.tex
actualizar manuscrito quarto

---
title: Shifts in water supply and demand drive land cover change across Chile
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    classoption:
      - numbered
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
  html:
    code-fold: true
    html-math-method: katex
  docx: default
# The `csl` style will be used with LaTeX-based rendering of`nature-pdf`
# if cite-method is `citeproc`. This is also the default if cite-method
# is not specified.
csl: _extensions/christopherkenny/nature/csl/springer-vancouver.csl
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.cl
    affiliations: 
        - name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad, Universidad Mayor
          id: 1
          city: Santiago
          country: 7500994, Chile.
        - id: 2
          name: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES), Universidad Mayor
          city: Santiago
          country: 7500994, Chile.
    attributes:
        corresponding: true
    #note: This is the first author footnote.
  - name: Anton Vrieling
    #email: a.vrieling@utwente.nl
    affiliations:
        - id: 3
          name: Faculty of Geo-Information Science and Earth, University of Twente
          city:  Enschede
          country: The Netherlands.
  - name: Francisco Meza
    affiliations:
        - id: 4
          name: Facultad de Agronomía y Sistemas Naturales, Pontificia Universidad Católica de Chile. 
          city: Santiago
          country: Chile.
        - id: 5
          name: Instituto para el Desarrollo Sustentable. Pontificia Universidad Católica de Chile
          city: Santiago
          country: Chile.
        - id: 6
          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile
          city: Santiago
          country: Chile.
  - name: Iongel Duran-Llacer
    affiliations:
        - id: 7
          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Universidad Mayor,
          city: Santiago
          country: 7500994, Chile.
  - name: Francisco Fernández
    #email: francisco.fernandez@uss.cl
    affiliations:
        - id: 8
          name: Center of Economics for Sustainable Development (CEDES), Faculty of Economics and Government, Universidad San Sebastian
          city: Santiago
          country: Chile.
        - id: 9
          name: Center of Applied Ecology and Sustainability (CAPES)
          city: Santiago
          country: Chile.
  - name: Alejandro Venegas-González
    affiliations:
        - id: 10
          name: Instituto de Ciencias Agroalimentarias, Animales y Ambientales (ICA3), Universidad de O'Higgins
          city: San Fernando
          country: Chile.
  - name: Nicolas Raab
    affiliations:
        - id: 4
  - name: Dylan Craven
    affiliations:
        - id: 11
          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750
          city: Santiago
          country: Chile.
        - id: 12
          name: Data Observatory Foundation
          city: Santiago
          country: Chile.
abstract: |
  Globally, droughts are becoming longer, more frequent, and more severe, and their impacts are multidimensional. These impacts typically extend beyond the water balance, as long-term, cumulative changes in the water balance can lead to regime shifts in land cover. Here, we assess the effects of temporal changes in water supply and demand over multiple time scales on vegetation productivity and land cover changes in continental Chile, which has experienced a severe drought since 2010. Over most of continental Chile, we have observed a persistent negative trend in water supply and a positive trend in atmospheric water demand since 2000. However, in water-limited zones, we have observed a negative trend in the vegetation's water demand. All the trends intensify over longer time scales. This long-term decrease in water availability has led to a decrease in vegetation productivity, especially for the Chilean matorral and the Valdivian temperate forest. Drought indices of soil moisture and actual evapotranspiration from short- to mid-term time scales primarily explain this decrease. Besides, our models indicate that the intensity of the drought explains up to 78% of the changes in shrubland land covers, decreasing to 40% for forests, but the change in cropland surface is explained by the surface burned.  Long-term climate change may cause regime shifts in land cover, which context-specific adaptation strategies can mitigate.
keywords: [drought, land cover, water demand, water supply]
bibliography: referencias.bib  
---

# Introduction {#sec-intro}

Across many regions of the world, droughts are becoming longer, more frequent, and more severe[@Miranda2023; @IPCC2023], impacting ecosystems via tree mortality[@Cheng2024], reducing vegetation productivity[@Miranda2023] and inducing shifts in land use and cover [@Crausbay2017]. However, identifying drought events is idiosyncratic due to the varying criteria used for classification. Droughts can be classified as 1) meteorological, i.e., when precipitation in a specific period falls below mean precipitation values observed over multiple years [@Mishra2010] (usually more than 30 years); 2) hydrological, i.e., when precipitation anomalies last for long periods (months to years) and affect the  hydrological system [@VanLoon2015;@VanLoon2016b] (e.g., streamflows, reservoirs and groundwater); 3) agricultural, i.e. when precipitation deficits negatively impact plant health, leading to decreases in crop or pasture productivity [@Wilhite1985]; or 4) ecological, i.e., when water availability negatively affect the provisioning of ecosystem services and trigger feedbacks in natural or human systems [@Crausbay2017]. Such feedbacks include drought impacts on human decision making and activities, which can lead to land-cover change [@VanLoon2016;@AghaKouchak2021], which may have cascading effects on biodiversity and ecosystem services [e.g., @Lawler2014;@Newbold2015]. Despite the high degree of confidence in the impacts of rising temperatures on the extent, frequency, and severity of agricultural and ecological droughts [@IPCC2023], which are likely to increase even if global warming stabilizes at 1.5°–2°C, the severity of meteorological droughts has been remarkably stable globally over the past century [@Vicente-Serrano2022;@Kogan2020]. A global study analyzing drought severity trends from 1980 to 2020 reveals that in a few regions (some mid-latitudinal and subtropical areas), rising temperatures during the warm season have increased atmospheric evaporative demand (AED), leading to an increase in agricultural land area [@Vicente-Serrano2022]. Thus, rising atmospheric water demand may reflect parallel changes in land cover—primarily agriculture—that can exacerbate the effects of meteorological droughts on ecosystems.

Expanding analyses to include multiple dimensions of droughts can provide complementary insights into the Earth’s water balance - and its impacts - over multiple time scales. Yet, the World Meteorological Organization recommends the use of a single drought index for monitoring droughts [@WMO2012], i.e., the multi-scale Standardized Precipitation Index (SPI; [@McKee1993]), which is limited in that it only considers  water supply in the form of precipitation. The Standardized Precipitation Evapotranspiration Index (SPEI; [@Vicente-Serrano2020]) builds upon SPI by incorporating the effects of  temperature on drought, and is now used widely for drought monitoring [e.g., @Gebrechorkos2023;@Liu2024]. Indices derived from soil moisture products [@Narasimhan2005; @Souza2021], such as the Standardized Soil Moisture Index [SSI, @AghaKouchak2014;@AghaKouchak2015] also monitor water supply and are thought to better capture water availability for crops, thus providing more relevant information for evaluating agricultural droughts. To disentangle the effects of precipitation from those of temperature [@Vicente-Serrano2020], as well as to capture droughts in terms of water atmospheric demand, AED has been integrated into the Evaporative Demand Drought Index [EDDI, @McEvoy2016], which is particularly effective at detecting the rapid onset or intensification of droughts. To quantify vegetation water demand, we must use the actual evapotranspiration, or the amount of water removed from a surface by evaporation and transpiration; the Standardized Evapotranspiration Index (SETI; [@Yang2019]) can be used for this purpose. In turn, ecological droughts, which capture the joint effects of precipitation and temperature in modifying natural and productive ecosystems [@Camps-Valls2021;@Paruelo2016;@Helman2014], are complex to measure and can therefore be monitored using multiple drought indices that capture the multiple dimensions of drought, e.g., precipitation, temperature, evapotranspiration, and AED. Although such an approach accounts for the joint effects of changes in natural and productive ecosystems, its potential impacts on land cover change have been largely unexplored.

From 1960 to 2019, land-use change has impacted approximately one-third of the Earth’s surface, which is four times more than previously thought [@Winkler2021]. Despite the considerable interest in land-use change dynamics [e.g., @Song2018;@Winkler2021], the direction and magnitude of  drought impacts on land cover change and vegetation productivity remain uncertain [@Chen2022;@Akinyemi2021;@Peng2017]. Meteorological droughts are responsible for approximately 37% of land cover change and variability in vegetation productivity globally [@Peng2017]. However, the evidence supporting these results is derived from only one drought index, SPEI, which combines a proxy for water supply -  precipitation - with a proxy for water demand - AED - at one time scale (12 months). The use of only one time scale may bias results of drought impacts towards ecosystems dominated by plant growth forms such as grasses and herbs that respond more rapidly to drought stress (< 12 months). This is because  physiological differences among and within dominant plant growth forms may increase (or decrease) tolerance of drought stress [@Craine2013;@McDowell2022]. For example, trees growing in more arid ecosystems typically respond over longer time scales than those in more humid ecosystems [@Vicente-Serrano2014]. Another source of uncertainty regarding drought impacts on land cover change and vegetation productivity are extrinsic factors, such as large-scale public policy (e.g., national and international reforestation initiatives), agricultural practices (e.g., clearing forest for soybean or oil palm), and rural and urban land use planning [@Karabulut2023].

To deepen current knowledge on the multidimensional impacts of drought on the temporal dynamics of natural and productive ecosystems, we evaluate temporal changes in water supply and demand, net primary productivity, and land-cover change across terrestrial ecosystems in continental Chile for 2000-2023. Chile’s diverse climate and ecosystems [@Beck2023;@Luebert2022] make it an ideal natural laboratory for assessing the dynamic interactions between climate and ecosystems, and potential impacts on land-cover change. Additionally, large parts of Chile have experienced severe drought conditions that have significantly affected vegetation and water storage in recent years; north-central Chile has faced a persistent precipitation deficit (or "mega-drought") since 2010 [@Garreaud2017], which has broadly impacted native forests [e.g., @Miranda2020;@UrrutiaJalabert2018;@Venegas2018] and agricultural productivity [e.g., @Zambrano2016;@Zambrano2018;@Zambrano2023]. However, the effects of this prolonged extreme drought may also extend to changes in land cover, altering the provision of key ecosystem services and agricultural production. Here, we aim to assess: short- to long-term time trends in multi-scalar drought indices that capture variation in the components of water balance, i.e., water supply (SPI, SPEI, SSI) and demand (EDDI, SETI) and their impacts on vegetation productivity and land-cover change across continental Chile. By doing so, we aim to deepen understanding of how water supply and demand affect vegetation productivity and influence changes in the surface of land cover in a severely drought-impacted region. This may enhance understanding of the alterations, due to drought, to ecosystems in various regions across the world.

# Materials and Methods

## Study area

Continental Chile has a diverse climate, with strong environmental gradients from north to south and east to west [@Aceituno2021] (Fig. 1a), which, together with its complex topography (Fig. 1b), determine its ecosystem diversity [@Garreaud2009;@Luebert2022] (Fig. 1c). We therefore divided Chile into ecoregions [@Dinerstein2017], which are regions that share similar geography and ecology, and have comparable levels of precipitation and solar radiation. There are seven ecoregions: Atacama desert, Central Andean dry puna, Southern Andean steppe, Chilean Matorral, Valdivian temperate forests, Magellanic subpolar forests, and Patagonian steppe. The Atacama desert is predominantly arid with hot (Bwh in the Koppen-Geiger classification) and cold (Bwk) temperatures, as well as the northern part of the Chilean Matorral. Most of the land in these two northern regions is bare, except for a small area where shrublands and grasslands are present. With an annual rainfall of less than 400 mm, the Central Andean dry puna ecoregion has low, yet highly seasonal precipitation with an eight-month dry season, low temperatures (Bwk) and is dominated by grasslands, shrublands, and savanna. The climate of the Southern Andean steppe ecoregion is cold desert (BWk), with most precipitation occurring in the winter. There is little vegetation in this ecoregion because the plants have adapted to its windy, dry, and cold climate. In central Chile, the climate of the Chilean Matorral  changes to that of an arid steppe with cold temperatures (Bsk). Then, towards the center-south of the country, the climate of the Chilean Matorral changes to a Mediterranean climate, with warm to hot summers (Csa and Csb).  Land cover in this ecoregion consists of a significant amount of shrublands and savannas. The Valdivian temperate forests have a mostly oceanic climate (Cfb) and a large area of forests and grasslands. The Magellanic subpolar forests have a tundra climate. Lastly, the Patagonian steppe has high aridity, cold temperatures (Bsk), and primarily consists of grasslands.

## Data

### Gridded meteorological and vegetation data

To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the MOD13A3 Collection 6.1 product derived from the MODIS (Moderate-Resolution Imaging Spectroradiometer) sensor onboard the Terra satellite. MOD13A3 provides vegetation indices with a 1 km spatial resolution and monthly frequency [@Didan2015]. We also utilized the MOD16A2 collection 6.1 [@Running2017] product from MODIS to gauge the water consumption of vegetation. This product gives us monthly actual evapotranspiration (ET) and atmospheric evaporative demand (AED) with a ~500m spatial resolution. For soil moisture, water supply, and water demand variables, we used ERA5-Land (ERA5L; ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides atmospheric and land variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3; Supplementary Materials and Methods, Supplementary Tables 2 and 4).

### Gridded indicators for land use

To account for the impacts of human activity on land cover change, we obtained data on road density [@Meijer2018] and night lights for the period 2012–2023 [@Roman2018]. These products are frequently used in the literature to quantify the human footprint (e.g., [@Halpern2015;@Halpern2008]) or biodiversity threats (e.g., [@Bowler2020;@Kennedy2020]).To capture changes on land cover due to fires, we calculated the total burned area for 2002-2023 [@Chen2023].

## Short- to long-term drought trends

### Atmospheric Evaporative Demand (AED)

To compute drought indices that use water demand, it is necessary to first calculate AED. To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:

$$AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}$$ {#eq-AED}

where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m. For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel as follow:

$$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-nature,
  numbered]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

%%%% Standard Packages

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

%%%%

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Shifts in water supply and demand drive land cover change across Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover, water demand, water supply},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Shifts in water supply and demand drive land cover change across
Chile]{Shifts in water supply and demand drive land cover change across
Chile}

% author setup
\author*[1,2]{\fnm{Francisco} \sur{Zambrano}}\email{francisco.zambrano@umayor.cl}\author[3]{\fnm{Anton} \sur{Vrieling}}\author[4,5,6]{\fnm{Francisco} \sur{Meza}}\author[7]{\fnm{Iongel} \sur{Duran-Llacer}}\author[8,9]{\fnm{Francisco} \sur{Fernández}}\author[10]{\fnm{Alejandro} \sur{Venegas-González}}\author[4]{\fnm{Nicolas} \sur{Raab}}\author[11,12]{\fnm{Dylan} \sur{Craven}}
% affil setup
\affil[1]{, \orgname{Hémera Centro de Observación de la Tierra, Facultad
de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad,
Universidad Mayor}}
\affil[2]{, \orgname{Observatorio de Sequía para la Agricultura y la
Biodiversidad de Chile (ODES), Universidad Mayor}}
\affil[3]{, \orgname{Faculty of Geo-Information Science and Earth,
University of Twente}}
\affil[4]{, \orgname{Facultad de Agronomía y Sistemas Naturales,
Pontificia Universidad Católica de Chile.}}
\affil[5]{, \orgname{Instituto para el Desarrollo Sustentable.
Pontificia Universidad Católica de Chile}}
\affil[6]{, \orgname{Centro Interdisciplinario de Cambio Global,
Pontificia Universidad Católica de Chile}}
\affil[7]{, \orgname{Hémera Centro de Observación de la Tierra, Facultad
de Ciencias, Universidad Mayor,}}
\affil[8]{, \orgname{Center of Economics for Sustainable Development
(CEDES), Faculty of Economics and Government, Universidad San
Sebastian}}
\affil[9]{, \orgname{Center of Applied Ecology and Sustainability
(CAPES)}}
\affil[10]{, \orgname{Instituto de Ciencias Agroalimentarias, Animales y
Ambientales (ICA3), Universidad de O'Higgins}}
\affil[4]{}
\affil[11]{\orgdiv{GEMA Center for Genomics, Ecology \& Environment,
Universidad Mayor, Camino La Pirámide Huechuraba 5750}}
\affil[12]{, \orgname{Data Observatory Foundation}}

% abstract

\abstract{Globally, droughts are becoming longer, more frequent, and
more severe, and their impacts are multidimensional. These impacts
typically extend beyond the water balance, as long-term, cumulative
changes in the water balance can lead to regime shifts in land cover.
Here, we assess the effects of temporal changes in water supply and
demand over multiple time scales on vegetation productivity and land
cover changes in continental Chile, which has experienced a severe
drought since 2010. Over most of continental Chile, we have observed a
persistent negative trend in water supply and a positive trend in
atmospheric water demand since 2000. However, in water-limited zones, we
have observed a negative trend in the vegetation's water demand. All the
trends intensify over longer time scales. This long-term decrease in
water availability has led to a decrease in vegetation productivity,
especially for the Chilean matorral and the Valdivian temperate forest.
Drought indices of soil moisture and actual evapotranspiration from
short- to mid-term time scales primarily explain this decrease. Besides,
our models indicate that the intensity of the drought explains up to
78\% of the changes in shrubland land covers, decreasing to 40\% for
forests, but the change in cropland surface is explained by the surface
burned. Long-term climate change may cause regime shifts in land cover,
which context-specific adaptation strategies can mitigate.}

% keywords
\keywords{drought,  land cover,  water demand,  water supply}

\begin{document}
\maketitle

\section{Introduction}\label{sec-intro}

Across many regions of the world, droughts are becoming longer, more
frequent, and more severe\citep{Miranda2023, IPCC2023}, impacting
ecosystems via tree mortality\citep{Cheng2024}, reducing vegetation
productivity\citep{Miranda2023} and inducing shifts in land use and
cover \citep{Crausbay2017}. However, identifying drought events is
idiosyncratic due to the varying criteria used for classification.
Droughts can be classified as 1) meteorological, i.e., when
precipitation in a specific period falls below mean precipitation values
observed over multiple years \citep{Mishra2010} (usually more than 30
years); 2) hydrological, i.e., when precipitation anomalies last for
long periods (months to years) and affect the hydrological system
\citep{VanLoon2015, VanLoon2016b} (e.g., streamflows, reservoirs and
groundwater); 3) agricultural, i.e.~when precipitation deficits
negatively impact plant health, leading to decreases in crop or pasture
productivity \citep{Wilhite1985}; or 4) ecological, i.e., when water
availability negatively affect the provisioning of ecosystem services
and trigger feedbacks in natural or human systems \citep{Crausbay2017}.
Such feedbacks include drought impacts on human decision making and
activities, which can lead to land-cover change
\citep{VanLoon2016, AghaKouchak2021}, which may have cascading effects
on biodiversity and ecosystem services
\citep[e.g.,][]{Lawler2014, Newbold2015}. Despite the high degree of
confidence in the impacts of rising temperatures on the extent,
frequency, and severity of agricultural and ecological droughts
\citep{IPCC2023}, which are likely to increase even if global warming
stabilizes at 1.5°--2°C, the severity of meteorological droughts has
been remarkably stable globally over the past century
\citep{Vicente-Serrano2022, Kogan2020}. A global study analyzing drought
severity trends from 1980 to 2020 reveals that in a few regions (some
mid-latitudinal and subtropical areas), rising temperatures during the
warm season have increased atmospheric evaporative demand (AED), leading
to an increase in agricultural land area \citep{Vicente-Serrano2022}.
Thus, rising atmospheric water demand may reflect parallel changes in
land cover---primarily agriculture---that can exacerbate the effects of
meteorological droughts on ecosystems.

Expanding analyses to include multiple dimensions of droughts can
provide complementary insights into the Earth's water balance - and its
impacts - over multiple time scales. Yet, the World Meteorological
Organization recommends the use of a single drought index for monitoring
droughts \citep{WMO2012}, i.e., the multi-scale Standardized
Precipitation Index (SPI; \citep{McKee1993}), which is limited in that
it only considers water supply in the form of precipitation. The
Standardized Precipitation Evapotranspiration Index (SPEI;
\citep{Vicente-Serrano2020}) builds upon SPI by incorporating the
effects of temperature on drought, and is now used widely for drought
monitoring \citep[e.g.,][]{Gebrechorkos2023, Liu2024}. Indices derived
from soil moisture products \citep{Narasimhan2005, Souza2021}, such as
the Standardized Soil Moisture Index
\citep[SSI,][]{AghaKouchak2014, AghaKouchak2015} also monitor water
supply and are thought to better capture water availability for crops,
thus providing more relevant information for evaluating agricultural
droughts. To disentangle the effects of precipitation from those of
temperature \citep{Vicente-Serrano2020}, as well as to capture droughts
in terms of water atmospheric demand, AED has been integrated into the
Evaporative Demand Drought Index \citep[EDDI,][]{McEvoy2016}, which is
particularly effective at detecting the rapid onset or intensification
of droughts. To quantify vegetation water demand, we must use the actual
evapotranspiration, or the amount of water removed from a surface by
evaporation and transpiration; the Standardized Evapotranspiration Index
(SETI; \citep{Yang2019}) can be used for this purpose. In turn,
ecological droughts, which capture the joint effects of precipitation
and temperature in modifying natural and productive ecosystems
\citep{Camps-Valls2021, Paruelo2016, Helman2014}, are complex to measure
and can therefore be monitored using multiple drought indices that
capture the multiple dimensions of drought, e.g., precipitation,
temperature, evapotranspiration, and AED. Although such an approach
accounts for the joint effects of changes in natural and productive
ecosystems, its potential impacts on land cover change have been largely
unexplored.

From 1960 to 2019, land-use change has impacted approximately one-third
of the Earth's surface, which is four times more than previously thought
\citep{Winkler2021}. Despite the considerable interest in land-use
change dynamics \citep[e.g.,][]{Song2018, Winkler2021}, the direction
and magnitude of drought impacts on land cover change and vegetation
productivity remain uncertain \citep{Chen2022, Akinyemi2021, Peng2017}.
Meteorological droughts are responsible for approximately 37\% of land
cover change and variability in vegetation productivity globally
\citep{Peng2017}. However, the evidence supporting these results is
derived from only one drought index, SPEI, which combines a proxy for
water supply - precipitation - with a proxy for water demand - AED - at
one time scale (12 months). The use of only one time scale may bias
results of drought impacts towards ecosystems dominated by plant growth
forms such as grasses and herbs that respond more rapidly to drought
stress (\textless{} 12 months). This is because physiological
differences among and within dominant plant growth forms may increase
(or decrease) tolerance of drought stress
\citep{Craine2013, McDowell2022}. For example, trees growing in more
arid ecosystems typically respond over longer time scales than those in
more humid ecosystems \citep{Vicente-Serrano2014}. Another source of
uncertainty regarding drought impacts on land cover change and
vegetation productivity are extrinsic factors, such as large-scale
public policy (e.g., national and international reforestation
initiatives), agricultural practices (e.g., clearing forest for soybean
or oil palm), and rural and urban land use planning
\citep{Karabulut2023}.

To deepen current knowledge on the multidimensional impacts of drought
on the temporal dynamics of natural and productive ecosystems, we
evaluate temporal changes in water supply and demand, net primary
productivity, and land-cover change across terrestrial ecosystems in
continental Chile for 2000-2023. Chile's diverse climate and ecosystems
\citep{Beck2023, Luebert2022} make it an ideal natural laboratory for
assessing the dynamic interactions between climate and ecosystems, and
potential impacts on land-cover change. Additionally, large parts of
Chile have experienced severe drought conditions that have significantly
affected vegetation and water storage in recent years; north-central
Chile has faced a persistent precipitation deficit (or ``mega-drought'')
since 2010 \citep{Garreaud2017}, which has broadly impacted native
forests \citep[e.g.,][]{Miranda2020, UrrutiaJalabert2018, Venegas2018}
and agricultural productivity
\citep[e.g.,][]{Zambrano2016, Zambrano2018, Zambrano2023}. However, the
effects of this prolonged extreme drought may also extend to changes in
land cover, altering the provision of key ecosystem services and
agricultural production. Here, we aim to assess: short- to long-term
time trends in multi-scalar drought indices that capture variation in
the components of water balance, i.e., water supply (SPI, SPEI, SSI) and
demand (EDDI, SETI) and their impacts on vegetation productivity and
land-cover change across continental Chile. By doing so, we aim to
deepen understanding of how water supply and demand affect vegetation
productivity and influence changes in the surface of land cover in a
severely drought-impacted region. This may enhance understanding of the
alterations, due to drought, to ecosystems in various regions across the
world.

\section{Materials and Methods}\label{materials-and-methods}

\subsection{Study area}\label{study-area}

Continental Chile has a diverse climate, with strong environmental
gradients from north to south and east to west \citep{Aceituno2021}
(Fig. 1a), which, together with its complex topography (Fig. 1b),
determine its ecosystem diversity \citep{Garreaud2009, Luebert2022}
(Fig. 1c). We therefore divided Chile into ecoregions
\citep{Dinerstein2017}, which are regions that share similar geography
and ecology, and have comparable levels of precipitation and solar
radiation. There are seven ecoregions: Atacama desert, Central Andean
dry puna, Southern Andean steppe, Chilean Matorral, Valdivian temperate
forests, Magellanic subpolar forests, and Patagonian steppe. The Atacama
desert is predominantly arid with hot (Bwh in the Koppen-Geiger
classification) and cold (Bwk) temperatures, as well as the northern
part of the Chilean Matorral. Most of the land in these two northern
regions is bare, except for a small area where shrublands and grasslands
are present. With an annual rainfall of less than 400 mm, the Central
Andean dry puna ecoregion has low, yet highly seasonal precipitation
with an eight-month dry season, low temperatures (Bwk) and is dominated
by grasslands, shrublands, and savanna. The climate of the Southern
Andean steppe ecoregion is cold desert (BWk), with most precipitation
occurring in the winter. There is little vegetation in this ecoregion
because the plants have adapted to its windy, dry, and cold climate. In
central Chile, the climate of the Chilean Matorral changes to that of an
arid steppe with cold temperatures (Bsk). Then, towards the center-south
of the country, the climate of the Chilean Matorral changes to a
Mediterranean climate, with warm to hot summers (Csa and Csb). Land
cover in this ecoregion consists of a significant amount of shrublands
and savannas. The Valdivian temperate forests have a mostly oceanic
climate (Cfb) and a large area of forests and grasslands. The Magellanic
subpolar forests have a tundra climate. Lastly, the Patagonian steppe
has high aridity, cold temperatures (Bsk), and primarily consists of
grasslands.

\subsection{Data}\label{data}

\subsubsection{Gridded meteorological and vegetation
data}\label{gridded-meteorological-and-vegetation-data}

To derive a proxy for vegetation productivity, we used the Normalized
Difference Vegetation Index (NDVI) from the MOD13A3 Collection 6.1
product derived from the MODIS (Moderate-Resolution Imaging
Spectroradiometer) sensor onboard the Terra satellite. MOD13A3 provides
vegetation indices with a 1 km spatial resolution and monthly frequency
\citep{Didan2015}. We also utilized the MOD16A2 collection 6.1
\citep{Running2017} product from MODIS to gauge the water consumption of
vegetation. This product gives us monthly actual evapotranspiration (ET)
and atmospheric evaporative demand (AED) with a \textasciitilde500m
spatial resolution. For soil moisture, water supply, and water demand
variables, we used ERA5-Land (ERA5L; ECMWF Reanalysis version 5 over
land) \citep{MunozSabater2021}, a reanalysis dataset that provides
atmospheric and land variables since 1950. It has a spatial resolution
of 0.1° (9 km), hourly frequency, and global coverage. We selected total
precipitation, maximum and minimum temperature at 2 meters, and
volumetric soil water layers between 0 and 100 cm of depth (layer 1 to
layer 3; Supplementary Materials and Methods, Supplementary Tables 2 and
4).

\subsubsection{Gridded indicators for land
use}\label{gridded-indicators-for-land-use}
